\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-28 paper/experiments.tex
paper: pass over experiments

\section{Experiments}\label{sec:experiments}
To evaluate the performance of our approach,  we implemented
a prototype system in Java, following a traditional iterator model.
For our experiments we plan and execute queries
for two separate survey-based data sets, showing that our system
is well suited for early dataset exploration.

@@

\subsubsection{freeCodeCamp 2016 New Coder Survey}
For our second set of experiments, we chose to use data collected
by freeCodeCamp (FCC) as a part of a survey of new software developers
(both professional and amateur)~\cite{fcc-data}. freeCodeCamp is an open-source
community that helps users learn how to program. Their \textit{2016 New Coder Survey} consists of responses by over 15,000 people to 48 different
demographic and programming-related questions.
The survey targeted users who were related to coding organizations.

We use a version of the data that has been pre-processed, but where missing values remain.
\todobox{Give example of expected missing value}{However, many of the missing values are expected because the data has been de-normalized.}
The original dataset has 15,620 rows and 113 attributes.
We chose a subset of 17 attributes, shown in~\Cref{table:fcc-description}, along with the percentage of missing values.

\begin{table}
  \centering
  \input{tables/fcc}
  \caption{Missing values in freeCodeCamp Survey Data}\label{table:fcc-description} 
\end{table}

@@

Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}). In~\Cref{q1}, we
calculate the average height for users based on their income data. In~\Cref{q2}, we compare
cholesterol levels for individuals with low, medium, and high incomes spectrum and above a
certain weight. In~\Cref{q3}, we extract the maximum blood lead levels for children under
the age of 6 years. In~\Cref{q4}, we calculate the average systolic blood pressure, by
gender, for subjects with a body mass index indicating obesity.

Our second set of queries is on the FCC data (\Cref{fig:queries-fcc}).
\Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
\Cref{q6} estimates the average age of women from the United States who participated.
\Cref{q7} calculates the average amount of money survey participants with student debt spent on learning based on their school degree.
\Cref{q8} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
The query calculates the average GDP per-capita of countries with and without bootcamp participants. 
\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on FCC data}
 \label{fig:queries-fcc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{query-acs}
 \caption{Queries on ACS data}
 \label{fig:query-acs}
 \end{subtable}

@@

\Cref{fig:runtimes} shows a summary of the performance results. The first plot
shows average running times for \ProjectName{} plans in two configurations:
quality-optimized and runtime-optimized. The second plot shows, for comparison, the average
running times required for each query when performing imputation at the base tables.  In all
cases, running a query through \ProjectName{} is cheaper than imputing a single base table,
with performance improvement on a factor of ten\todo{give range of performance}. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined.png}
\caption{Query runtimes with imputation on base tables, \ProjectName{} imputation
    optimizing for quality ($\alpha=0$), and \ProjectName{} imputation optimizing for
    efficiency ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
    imputation can provide another order-of-magnitude speedup at the expense of quality. For
    relatively large queries (Query 9), runtime of imputation can be reduced from 6.5
    hours to 7 seconds. }
\label{fig:runtimes}
\end{figure}

@@

In many real-world cases, applying the imputation step at the base table is prohibitively
expensive.
To illustrate the increasing difficulty of such an approach as datasets scale, we run the
query in (\Cref{fig:query-acs}) over the ACS dataset.
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 355 minutes(\ref{fig:runtimes}). In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.

>>>>

\section{Experiments}\label{sec:experiments}
To evaluate the performance of our approach,  we implemented
a prototype system in Java, following a traditional iterator model.
For our experiments we plan and execute queries
for three separate survey-based data sets, showing that our system
is well suited for early dataset exploration.

@@

\subsubsection{freeCodeCamp 2016 New Coder Survey}
For our second set of experiments, we chose to use data collected
by freeCodeCamp (fCC) as a part of a survey of new software developers
(both professional and amateur)~\cite{fcc-data}. freeCodeCamp is an open-source
community that helps users learn how to program. Their \textit{2016 New Coder Survey} consists of responses by over 15,000 people to 48 different
demographic and programming-related questions.
The survey targeted users who were related to coding organizations.

We use a version of the data that has been pre-processed, but where missing values remain.
For example, 46.6\% of entries for entry \textit{commutetime} are missing. However, it is worth noting that some of the missing values
are also expected, given the way the data has been de-normalized. For example, \textit{bootcamploanyesno}, a binary attribute encoding
whether a respondent had a loan for a bootcamp, is expected to be null for participants who did not attend a bootcamp.

We chose a subset of 17 attributes, shown in~\Cref{table:fcc-description}, along with the percentage of missing values.

\begin{table}
  \centering
  \input{tables/fcc}
  \caption{Missing value distribution for each attribute in freeCodeCamp Survey Data}\label{table:fcc-description} 
\end{table}

@@

Our first set of queries is on the CDC data (\Cref{fig:queries-cdc}). In~\Cref{q1}, we
calculate the average cuff size for users based on their income data. In~\Cref{q2}, we compare
creatine levels for individuals with low, medium, and high incomes spectrum and above a
certain weight. In~\Cref{q3}, we extract the average blood lead levels for children under
the age of 6 years. In~\Cref{q4}, we calculate the average systolic blood pressure, by
gender, for subjects with a body mass index indicating obesity.

Our second set of queries is on the fCC data (\Cref{fig:queries-fcc}).
\Cref{q5} calculates the average income for survey participants, based on their bootcamp attendance.
\Cref{q6} estimates the average commute time of women from the United States who participated.
\Cref{q7} calculates the average amount of student debt based on school degree for survey participants who have student debt.
\Cref{q8} joins the freeCodeCamp data with a reference table provided by the World Bank which summarizes GDP per-capita across various countries~\cite{worldbank-data}.
The query calculates the average GDP per-capita of countries with and without bootcamp participants. 
\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on fCC data}
 \label{fig:queries-fcc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{query-acs}
 \caption{Queries on ACS data}
 \label{fig:query-acs}
 \end{subtable}

@@

\Cref{fig:runtimes} shows a summary of the performance results. We evaluated three different
configurations for each query. We compare \ProjectName{} optimizing for quality ($\alpha=0$),
\ProjectName{} optimizing for performance($\alpha=1$), and imputing at the base tables followed
by executing the query traditionally. The quality-optimized queries are an order-of-magnitude
faster than imputing on the base tables. We can get another order-of-magnitude speedup when
optimizing for performance. This performance differential means it is feasible
to explore multiple imputations operations (including more expensive operators) when using
\ProjectName{}, in contrast to the traditional approach of base table imputation.

\begin{figure}
\includegraphics[width=\columnwidth]{figures/running_times_combined.png}
\caption{Query runtimes with imputation on base tables, \ProjectName{}
    optimizing for quality ($\alpha=0$), and \ProjectName{} optimizing for
    performance ($\alpha=1$). Quality-optimized, on-the-fly imputation provides an order of
    magnitude speedup over imputing on base tables. Efficiency-optimized on-the-fly
    imputation can provide another order-of-magnitude speedup at the expense of quality. For
    relatively expensive queries over large datasets (Query 9), runtime of imputation can be reduced from 6.5
    hours to 7 seconds. }
\label{fig:runtimes}
\end{figure}

@@

In many real-world cases, applying the imputation step at the base table is prohibitively
expensive.
To illustrate the increasing difficulty of such an approach as datasets scale, we run~\Cref{q9} over the ACS dataset.
Applying the imputation operation to the base table is extremely expensive, in our case
completing in 355 minutes(\Cref{fig:runtimes})\todo{Micah: this doesn't match the 6.5 hours in the figure}. In contrast, \ProjectName{} executes a quality-optimized version
in 7 seconds and a runtime-optimized version in 1 second. This highlights the potential
benefit of using our system for early data exploration.
